---
title: Examining the mental health impacts of Universal Credit, treating phased rollout
  as a natural experiment
author: "Andy Baxter, Sophie Wickham, Mandy Cheetham, Emma Coombes, Clare Bambra,
  David Taylor-Robinson, Luke Munford, Huasheng Xiang, Mattheo Richiardi, Heather
  Brown, Matthew Sutton, Benjamin Barr, Vittal Katikireddi, Peter Craig"
date: "`r Sys.Date()`"
output: 
  word_document:
    reference_docx: template_doc.docx
bibliography: C:/andy_local/BibTeX files/APS_DiD_paper_files.bib
csl: http://csl.mendeley.com/styles/481764751/harvard-cite-them-right
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE, fig.height=8.0, fig.width=12.0, compression="lzw", dev="tiff", dpi=500, out.width=900)
```

# Abstract

#### Background

Universal Credit was rolled out across the UK as a benefit system to
replace six existing benefits. It aimed to consolidate and simplify
benefit claim systems and encourage more claimants into work. Key
changes from legacy benefit systems may be harmful to the mental health
of claimants, or conversely, the push towards work may have beneficial
effects. We aimed to test the causal effects of Universal Credit
compared to legacy benefit systems, treating the phased rollout as a
natural experiment.

#### Methods

We used difference-in-differences methods to test for changes in mental
health and well-being using the measures of life satisfaction,
happiness, anxiety and whether respondents felt their lives were
worthwhile reported in the Annual Population Survey. We compared
as-yet-untreated areas with UC-rollout areas by month across the natural
migration phase. We conducted subgroup analyses to test for differential
harms by claimant types and other vulnerabilities.

#### Findings

#### Interpretation

#### Funding

This research is supported by the UK National Institute of Health
Research Public Health Research Programme, grant number NIHR131709.

\newpage

# Introduction

Universal Credit (UC) was introduced as a new, replacement benefit
system, combining multiple prior benefits into a single application and
award scheme. Rollout began in 2013 and aimed to gradually migrate all
claimants of a 'legacy benefit' -- Job Seekers Allowance, Working Tax
Credit, Child Tax Credit, Housing Benefit, Income Support Allowance, or
Income-based Employment Support Allowance -- to the new claims system
[@NationalAuditOffice2018; @DepartmentforWork&Pensions2022a]. The UC
system, in comparison with the legacy systems, is designed as a fully
online application and claim-handling system. It aims to encourage more
claimants into work by setting eligibility criteria, introducing a
single 'taper' on awards as income increases, and making payments in
lieu of each claim month, mimicking salary payments [@Wickham2020;
@DepartmentforWork&Pensions2022a; @DepartmentforWork&Pensions2022].

The change to Universal Credit may have negatively affected the health
and wellbeing of some benefit recipients [@Wickham2020; @Brewer2022].
Waits for first payments -- with the minimum 5-week assessment period
plus administrative delays resulting in up to a 12-week wait
[@NationalAuditOffice2018; @Cheetham2019] -- may be a potential cause of
distress in low-income households. Incurred debt and ongoing struggles
to keep up with payments may prolong this effect beyond the initial
months [@DepartmentforWork&Pensions2018]. Increased work-search
requirements might similarly impact mental health across the term of
benefit receipt, potentially requiring action beyond claimants'
capabilities or being perceived as 'hostile' and 'uncaring' in demanding
justification for inaction [@Cheetham2019;
@DepartmentforWork&Pensions2023]. The switch to fully-online claims
system, perceived as "complicated, disorientating, impersonal, hostile
and demeaning" [@Cheetham2019], may additionally have contributed to
worsening mental health. Conversely, the aim to increased employment of
benefit claimants may lead to improved mental health outcomes, both
through support in applying for jobs and in less punitive reductions in
benefits with increases in earned income
[@DepartmentforWork&Pensions2022a].

Treating the phased rollout as a stepped-wedge natural experiment, the
causal effects of UC on mental health may be estimable using
as-yet-unexposed areas as controls. Natural experimental methods divide
populations into exposed and unexposed comparison groups using events
outside of the control of the researcher [@Craig2017]. Using the planned
UC 'natural migration' rollout timetable, potential benefit claimants
can be classified as 'exposed' and 'unexposed' by determining whether UC
had, at date of data collection, been implemented in the local authority
area or not. Quantifying these effects would give a clearer picture of
how much observed differences in mental health are attributable to the
switch to a new benefit system and give an estimate of whether UC causes
overall benefits or harms to claimants. Subgroup analyses of claimant
types and other vulnerabilities can also identify at-risk groups and
areas for preventative action. These findings could inform future
changes to benefit systems to improve mental health outcomes.

This study is part of Work Package 1 of an NIHR-funded study on
evaluating the mental health impacts of Universal Credit grant number
NIHR131709; @Craig2022].

# Methods

## Aim and research questions

We aimed to estimate the Average Treatment Effects across Local
Authority areas of Universal Credit on the mental health of people
eligible to claim benefits. Additionally, we aimed to test differences
in effect sizes across markers of vulnerability.

### Research questions

1.  What is the impact of UC rollout on the mental health and wellbeing
    of working age adults?
2.  Do effects vary across sex, age of claimant, pre-UC employment
    status, household structure, disability and health status, and
    country?
3.  How does this effect vary in the time periods immediately before and
    immediately following a switch to Universal Credit? Are there
    anticipatory effects in the immediate pre-rollout period? Do
    post-exposure effects change as time from rollout increases?

## Protocol

The analysis plan, along with code used in analysis, is archived at
[osf.io/knajb](https://osf.io/knajb) [@Baxter2022].

## Data

We used data from the Annual Population Survey (APS) secure-access
version collected between 2013-2019
[@OfficeforNationalStatistics-SocialSurveyDivision2022]. These data are
released as annual cross-sectional datasets, produced from the Labour
Force Survey data collection [@OfficeforNationalStatistics2012]. APS
responses record demographics, household and employment conditions,
benefit receipt and the 'ONS4' well-being outcomes ['Life Satisfaction',
'Worthwhile', 'Happiness' and 'Anxiety';
@OfficeforNationalStatistics2018]. Data are collected using area-based
quotas and record each observation's area of residence; this allows
grouping of observations by Local Authority District, which match the
area-wise Universal Credit rollouts announced for the 2015--2018 period
[@DepartmentforWork&Pensions2018a].

## Study design

We used natural experimental methods to estimate the effects of the
rollout of Universal Credit on mental health across all claimants and
potential claimants. We treated the differences in treatment timing as a
'target trial', evaluating how closely the phased rollout emulates a
stepped wedge randomised controlled trial to assess the potential for
bias in these comparisons (Hernán and Robins, 2016; Craig et al., 2019;
Matthews et al., 2022). We determined the eligible population using two
methods -- observation of any benefit receipt at any time point, and a
logistic regression prediction of UC uptake under a simulated
full-UC-rollout scenario using UKMOD (Centre for Microsimulation and
Policy Analysis, 2022). We conducted a staggered
difference-in-differences analysis, treating the UC 'natural migration'
phased rollout period as a marker defining exposed and unexposed
comparators (Callaway and Sant'Anna, 2021; Goodman-Bacon, 2021; Sun and
Abraham, 2021). We compared changes in mental health in local authority
areas in which Universal Credit had been introduced with changes in
mental health in local authority areas in which Universal Credit had not
yet been introduced. We explored differences in effects across subgroup
characteristics marking potential differences in vulnerability to
benefit changes and differences in reasons for claiming universal
credit.

## Determining the affected population

To determine comparable populations of potential UC claimants across the
observed years for inclusion, we used two definitions of population of
interest, one using empirical methods of determining exposure and one
using prediction methods.

Under the empirical definition, the population of interest included all
respondents who report claiming either a legacy benefit (Job Seeker's
Allowance, Tax Credits, Housing Benefit, Income support, Income-based
Employment and Support Allowance) or Universal Credit at any time point.
A limitation of this method is that a potential claimant under one
system may be ineligible (or have been disincentivised from applying)
under the other. Under the intended outputs of UC in encouraging
claimants into work, a person whose circumstances have been thus
affected by the UC rollout might not be an observed claimant in the
exposed period. Matching cross-sectional populations by observed benefit
receipt may not create exchangeable populations. Additionally, benefit
receipt may be differentially underreported under one or both systems.
The alternative approach of using observed characteristics as
straightforward markers of eligibility (for example, including all
unemployed respondents) is hampered by missing data on housing costs,
savings and partner's income which determine eligibility and payment
amount.

To overcome these potential biases, we used a machine-learning model
trained on simulated data to predict the probability of a person
receiving UC under a full-UC-rollout scenario. We used the UKMOD static
tax-benefit microsimulation model to simulate benefit receipt for a
donor population (from the Family Resources Survey) under a
full-UC-rollout scenario from 2014 to 2019. We trained a model to
predict probability of receipt of Universal Credit from a set of
individual and household characteristics which were recorded in both the
Family Resources Survey dataset and the APS dataset. We used this model
to calculate a probability score for each observation in the APS dataset
for receiving UC in the hypothetical scenario of exclusive and full UC
availability across all years. We assigned this probability as a weight
to each respondent to have a comparably weighted population across all
years, giving greatest weight to those who were most likely to be
affected by the changes in benefit systems.

## Exposure

The replacement of legacy benefits with UC broadly took place in three
phases [@DepartmentforWork&Pensions2018a;
@DepartmentforWork&Pensions2022a; @NationalAuditOffice2018]:

-   The restricted rollout of the 'live service' from 2013 to a limited
    subgroup, mainly single, childless claimants.
-   The full-service 'natural migration' period from 2016 onwards --
    open to all claimant types but only applied to new claimants,
    existing claimants with changing circumstances necessitating a
    change in claim, and voluntary benefit switchers.
-   The concluding 'managed migration' phase, which was piloted in
    2019-2020 but paused for the pandemic and restarted in 2022,
    implementing a forced migration of all claimants to UC by 2024.

Observing the whole eligible population, we proposed using the natural
migration dates as markers of exposure to Universal Credit if the
respondent lived in an area already exposed at the time of data
collection. The dates by Job Centre Plus reported in the 'Universal
Credit transition rollout schedule March 2018 to December 2018'
(Department for Work & Pensions, 2018b) were used to determine a first
exposed date for each Local Authority. These were matched to APS
observations by Local authority district. Each observation was assigned
a '1' for exposure if the reference date for that interview matched or
followed the month of rollout and a '0' otherwise. Months from rollout
were additionally assigned to each, individual, counting upwards from
negative months to '0' in the month of planned natural migration in that
district and a positive count of months following.

## Outcomes

The four wellbeing outcomes (Table 1) were recorded for each
observation.

| Label             | Question in survey                                                                        | Possible answers (on a scale of 0--10)                        |
|----------------------|-------------------------|-------------------------|
| Life Satisfaction | "Overall, how satisfied are you with your life nowadays?"                                 | 0 -- 'not at all satisfied' to 10 -- 'completely satisfied'   |
| Worthwhile        | "Overall, to what extent do you feel that the things you do in your life are worthwhile?" | 0 -- 'not at all worthwhile' to 10 -- 'completely worthwhile' |
| Happiness         | "How happy did you feel yesterday?"                                                       | 0 -- 'not at all happy' to 10 -- 'completely happy'           |
| Anxiety           | "How anxious did you feel yesterday?"                                                     | 0 -- 'not at all anxious' to 10 -- 'completely anxious'       |

Table 1 -- The four questions recording subjective wellbeing, recorded
in the Annual Population Survey in data collected from April 2011.

## Missing data

We used multiple imputation methods to overcome potential biases due to
missing data in our primary analyses. Data of interest was missing for
x% of respondents. The variable with the highest rate of missingness was
earned income, with data missing for x% of eligible respondents. We
generated [20] imputed datasets.

We additionally carried out a 'complete case' analysis of all
observations with complete data. We compared effect estimates between
complete and imputed samples to examine robustness of findings.

## Statistical analysis

We used difference-in-differences methods to estimate the effects of the
introduction of Universal Credit on mental health and wellbeing. Two-way
fixed-effects (TWFE) estimates compare exposed with non-yet-exposed
units can ensure comparison of exchangeable units. However, in cases of
varying treatment timing and effect size this can introduce error in
estimating magnitude and direction of outcome (Goodman-Bacon, 2021; Roth
et al., 2022). We ran TWFE models and examined effect estimate by
treatment timing using decomposition to test for bias due to
heterogeneous treatment effects. In further models, to overcome
anticipated bias, we used the R package 'did2s' (Butts and Gardner,
2021) to implement the methods of Callaway and Sant'Anna (2021) and
'Two-stage differences in differences' approach by Gardner (2021) as two
comparable alternatives. We used each method to estimate a static effect
estimate across the four quarters following the rollout date in each
Local Authority. We further used each method to produce a coefficient
and standard error term for each quarterly period before and after
introduction of Universal Credit to examine timings and patters of
effects.

# Graphing conditions across exposure period

```{r claim_patterns}
library(fst)
library(data.table)
library(tidyverse)
library(dtplyr)
library(lubridate)
library(SPHSUgraphs)
library(patchwork)

theme_set(theme_sphsu_light())

here::i_am("reports/outline_paper_parts.Rmd")

apr14_mar21 <- c(
  here::here("data", "fst_files", "apr14_mar15.fst"),
  here::here("data", "fst_files", "apr15_mar16.fst"),
  here::here("data", "fst_files", "apr16_mar17.fst"),
  here::here("data", "fst_files", "apr17_mar18.fst"),
  here::here("data", "fst_files", "apr18_mar19.fst"),
  here::here("data", "fst_files", "apr19_mar20.fst"),
  here::here("data", "fst_files", "apr20_mar21.fst")
) |> map(read_fst, c(
  "age",
  "tpbn1301", 
  "tpbn1302", 
  "tpbn1303", 
  "tpbn1304", 
  "tpbn1305", 
  "tpbn1306", 
  "tpbn1307", 
  "tpbn1308", 
  "tpbn1309", 
  "tpbn1310", 
  "refdte"))

wts1 <- c(
  here::here("data", "fst_files", "apr14_mar15.fst"),
  here::here("data", "fst_files", "apr15_mar16.fst"),
  here::here("data", "fst_files", "apr16_mar17.fst"),
  here::here("data", "fst_files", "apr17_mar18.fst"),
  here::here("data", "fst_files", "apr18_mar19.fst")
  # "data/fst_files/apr19_mar20.fst"
  # "data/fst_files/apr20_mar21.fst"
) |> map(read_fst, "pwta18") |> 
  map(`colnames<-`, "weight")

wts2 <- c(here::here("data", "fst_files", "apr19_mar20.fst"),
          here::here("data", "fst_files", "apr20_mar21.fst")) |>
  map(read_fst, "pwta20") |> 
  map(`colnames<-`, "weight")


n_uc <- apr14_mar21 |> 
  map2(append(wts1, wts2), cbind) |> 
  reduce(bind_rows) |> 
  mutate(date = floor_date(dmy(refdte), "months"),
         across(tpbn1301:tpbn1310, ~.x == 1)) |> 
  rowwise() |> 
         mutate(uc = sum(tpbn1301:tpbn1310)*weight)

apr14_mar21_dt <- apr14_mar21 |> 
  map2(append(wts1, wts2), cbind) |> 
  reduce(bind_rows) |> 
  mutate(date = floor_date(dmy(refdte), "months")) |> 
  data.table()

apr14_mar21_dt <- apr14_mar21_dt[age > 17 & age < 66]

apr14_mar21_dt[,uc := rowSums(.SD == 1), .SD = 2:11]
apr14_mar21_dt[,tax_cr := rowSums(.SD == 3), .SD = 2:11]
apr14_mar21_dt[,housing := rowSums(.SD == 2), .SD = 2:11]
apr14_mar21_dt[,income_s := rowSums(.SD == 4), .SD = 2:11]
apr14_mar21_dt[,jsa := rowSums(.SD == 5), .SD = 2:11]
apr14_mar21_dt[,disab := rowSums(.SD == 6), .SD = 2:11]
apr14_mar21_dt[,other := as.numeric(rowSums(.SD)>0), .SD = tax_cr:jsa]

perc_other <- apr14_mar21_dt |> 
  as_tibble() |> 
  group_by(date) |>
  select(date, other, weight) |> 
  summarise(prop_other = sum(other * weight)/sum(weight)) |> 
  ggplot(aes(date, prop_other)) +
  geom_col(fill = sphsu_cols("Pumpkin"), width = 32) +
  scale_x_date("Year") +
  scale_y_continuous("Respondents on\nLegacy Benefits", labels = scales::percent,
                     expand = expansion(mult = c(0, 0.05)))

perc_all <- apr14_mar21_dt |> 
  as_tibble() |> 
  group_by(date) |>
  select(date, uc, weight) |> 
  summarise(prop_uc = sum(uc * weight)/sum(weight)) |> 
  ggplot(aes(date, prop_uc)) +
  geom_col(fill = sphsu_cols("University Blue"), width = 32, position = "identity") +
  scale_x_date("Year") +
  scale_y_continuous("Respondents on UC", labels = scales::percent,
                     expand = expansion(mult = c(0, 0.05)))

perc_claim <- apr14_mar21_dt |> 
  group_by(date) |>
  select(date, uc, other, weight) |> 
  filter(uc == 1 | other == 1) |> 
  mutate(benefits = case_when(
    uc == 1 & other == 1 ~ "Combination",
    uc == 1 ~ "Universal Credit",
    other == 1 ~ "Legacy Benefits"
  ),
  benefits = factor(benefits, levels = c("Legacy Benefits", "Combination", "Universal Credit"))) |> 
  as_tibble() |> 
    # group_by(date, benefits) |> 
    # summarise(n = sum(weight)) |> 
  ggplot(aes(date, weight, fill = benefits)) +
  # geom_col(aes(y = weight), width = 32, position = "fill") +
    stat_summary(fun = sum, position = "fill", geom = "bar", width = 32) +
  # stat_sum(aes(y = weight), position = "fill", geom = "bar", width = 31) +
  # stat_count(position = "fill", geom = "bar", aes(y = after_stat(count)), width = 31, position = "identity") +
  scale_fill_manual(name = "Benefits claimed", values = sphsu_cols("Pumpkin", "Leaf", "University Blue", names = FALSE)) +
  scale_x_date("Year") +
  theme(legend.position = "bottom") +
    guides(size = guide_none()) +
  scale_y_continuous("Claimant by benefit types", labels = scales::percent,
                     expand = expansion(mult = c(0, 0.05)))


patch <- (perc_other/perc_all/perc_claim) & theme(panel.grid.major.x = element_blank(),
                              panel.grid.minor.x = element_blank()) 
patch +
  plot_annotation(title = "Proportions of working-age adults claiming benefit types",
                  subtitle = "April 2014 - March 2021")
```

\newpage

# References
